Appendix 4: Results

### 2016-02-11 Harris_2015/manuscript-materials/draft.tex
editor comments

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can also be estimated
from observed co-occurrence rates via maximum likelihood, controlling
for indirect effects. Using simulated landscapes with known pairwise
interaction strengths, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods,
correctly isolating direct interactions between species pairs even when
indirect interactions or abiotic factors largely overpowered them. Two
computationally efficient approximations, based on controlling for
indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling
approach to produce incorrect inferences, however.

@@

\noindent
To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Connor et al. 2013), existing methods
for detecting competition's effects on community structure are
unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
one another can become lost in a web of indirect effects. For example,
the competitive interaction between the two shrub species in Figure 1A
is obscured by their shared tendency to occur in unshaded areas (Figure
1B). While ecologists have long known that indirect effects can
overwhelm direct ones at the landscape level (Levine 1976), the vast
majority of our methods for drawing inferences from observational data
do not control for these effects (e.g. Diamond 1975, Strong et al. 1984,
Gotelli and Ulrich 2009, Veech 2013, Pollock et al. 2014). To the extent
that indirect interactions like those in Figure 1 are generally
important, existing methods will not provide much evidence regarding
species interactions.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it does still leave a
signal in the data (Figure 1C). For example, after partitioning the data
set into shaded and unshaded sites, there will be co-occurrence deficits
in each subset that wouldn't otherwise be apparent. More generally,
controlling for other species in the network will often be important for
obtaining reliable estimates of direct (conditional, or all-else-equal)
effects. This kind of precision is difficult to obtain from null models,
which only test the most extreme possible hypothesis: that \emph{all}
direct and indirect interactions are exactly zero. Nevertheless, null
models have dominated this field for more than three decades (Strong et
al. 1984, Gotelli and Ulrich 2009).

Following Azaele et al. (2010), this paper shows that Markov networks
(undirected graphical models also known as Markov random fields; Murphy
2012) can provide a framework for understanding the landscape-level
consequences of pairwise species interactions, and for estimating them
from observed presence-absence matrices. Markov networks have been used
in many scientific fields in similar contexts for decades, from physics
(where nearby particles interact magnetically; Cipra 1987) to spatial
statistics (where adjacent grid cells have correlated values; Harris
1974, Gelfand et al. 2005). While community ecologists explored some
related approaches in the 1980's (Whittam and Siegel-Causey 1981), they
used severe approximations that led to unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982).

Below, I demonstrate Markov networks' ability to produce exact
predictions about the direct and indirect consequences of an interaction
matrix, and also to make inferences about the species interactions that
contributed to an observed set of co-occurrences. Using simulated data
sets where the ``true'' interactions are known, I compare this approach
with several existing methods. Finally, I discuss opportunities for
extending the approach presented here to other problems in community
ecology, e.g.~quantifying the overall effect of species interactions on
occurrence rates (Roughgarden 1983) and disentangling the effects of
biotic versus abiotic interactions on species composition (Pollock et
al. 2014).

@@

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if \(\beta_{ij} = +2\), then each species' odds of
occurrence would be \(e^2\) times higher when the other one is present
(as compared with otherwise equivalent sites). The relative probability
of a presence-absence vector increases when positively-associated
species co-occur and decreases when negatively-associated species
co-occur. As a result, the model tends---all else equal---to produce
assemblages where many positively-associated species pairs co-occur and
few negatively-associated pairs do (just as an ecologist might expect).
When all else is \emph{not} equal (e.g.~Figure 1, where the presence of
one competitor is associated with release from another competitor), then
predicting species' overall co-occurrence rates can be more complicated,
and may require summing over the different possible assemblages (Figure
2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When the number of species is
reasonably small (less than about 30), one can find exact maximum
likelihood estimates for all of the \(\alpha\) and \(\beta\)
coefficients given a presence-absence matrix by numerically optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will converge on the global maximum. I used
the rosalia package (Harris 2015a) for the R programming language (R
Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
gradients (Murphy 2012); the package passes these functions to the
``BFGS'' method in R's general-purpose optimizer, which then estimates
the Markov network parameters.

@@

I then generated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model; see Appendix 2 for
details). I randomly drew the ``true'' coefficient magnitudes for each
replicate landscape from exponential distributions so that most species
pairs interacted negligibly but a few pairs interacted strongly enough
that their effects could propagate indirectly to other species in the
network.

The first set of 20-species landscapes, like the landscapes with three
species, were generated directly from a Markov network to ensure that
the model could recover the parameters used to generate the ``observed''
co-occurrence data. Then, I added two environmental factors that varied
from location to location across the simulated landscapes, and simulated
a new set of co-occurrence data so that species' \(\alpha\) coefficients
depended on the local environment. The latter set of simulated
landscapes provide an important test of the methods' ability to
distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces
like abiotic environmental filtering. This task was made especially
difficult because---as with most analyses of presence-absence data for
co-occurrence patterns---the inference procedure did not have access to
any information about the environmental or spatial variables that helped
shape the landscape (cf Connor et al. 2013). I generated the final set
of landscapes with an abundance-based model that included per-capita
interaction rates instead of per-species interaction rates.

@@

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships. This approach is common in molecular biology
(Friedman et al. 2008), but is rare in ecology (see Albrecht and Gotelli
(2001) and Faisal et al. (2010) for two exceptions). In the context of
non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
(Loh and Wainwright 2013). Because partial correlations are undefined
for landscapes with perfectly-correlated species pairs, I used a
regularized estimate provided by the corpcor package's
\texttt{pcor.shrink} function with the default settings (Schäfer et al.
2014).

@@

The next method, described in Gotelli and Ulrich (2009), involved
simulating new landscapes from a null model that retains the row and
column sums of the original matrix (Strong et al. 1984). I used the
\(Z\)-scores computed by the Pairs software described in Gotelli and
Ulrich (2009) as my null model-based estimator of species interactions.

@@

For the larger landscapes, I evaluated the relationship between each
method's estimates and the ``true'' interaction strengths. To ensure
that the different test statistics (e.g.~correlations versus \(Z\)
scores) were on a common scale, I rescaled them using linear regression
through the origin. I then calculated the proportion of variance
explained for different combinations of model type and landscape size
(compared with a null model that assumed all interaction strengths to be
zero).

@@

\noindent
\textbf{Twenty species.} In general, each model's performance was
highest for large landscapes with simple assembly rules and no
environmental heterogeneity (Figure 3). Despite some variability across
contexts, the rank ordering across methods was very consistent. In
particular, the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did
not. The Markov network consistently performed best of all. As
anticipated by Lee and Hastie (2012), generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when
the data sets were very large (Figure 3). As reviewed in Gotelli and
Ulrich (2009), however, most analyses in this field of ecology involve
fewer than 50 sites, and the gap between the Markov network and GLMs is
larger in this context. As shown in Appendix 4, the standard errors
associated with the estimates in Figure 3 are small (less than 0.01), so
the differences among methods should not be attributed to sampling
error.

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). The estimates
from these approaches were tightly correlated (after controlling for the
size of the landscape) suggesting that the null model only contains a
noisy version of the same information that could be obtained more easily
and interpretably with simple correlation coefficients (Figure 4B).

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, for which the true interaction rates are
easiest to interpret). The Markov network's approximate Type I error
rate (defined here as the probability that 0 fell outside the 95\%
confidence interval for a pair of species where \(|\beta_{ij}|<0.1\))
depended on the simulation type: 0.02 for simulations that matched the
model's assumptions, versus 0.14 for simulations that included
environmental heterogeneity (see Appendix 4 for confidence interval
coverage across a range of \(\beta_{ij}\) values). The null model's Type
I error rates were 0.30 and 0.50 for the constant and heterogeneous
landscapes, respectively---far higher than the nominal 0.05 rate. Figure
4C shows, across a range of true interaction strengths, the probability
that the null model or the Markov network will predict the wrong sign of
the interaction with 95\% confidence. The null model makes such errors
more than 5 times as often as the Markov network, even though it only
reject the null hypothesis twice as often overall (Appendix 4). The
Markov network's errors were also more concentrated around 0, so it
never misclassified strong interactions like the null model did.

@@

\setlength{\parindent}{-1em} \setlength{\leftskip}{1em}

@@

\hyperdef{}{ref-xiaoux5fstrongux5f2015}{\label{ref-xiaoux5fstrongux5f2015}}
Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.

\setlength{\parindent}{0cm} \setlength{\leftskip}{0cm}

\noindent \textbf{Figures:}

\textbf{Figure 1: A.} A small network of three competing species. The
tree (top) tends not to co-occur with either of the two shrub species,
as indicated by the strongly negative coefficient linking them. The two
shrub species also compete with one another, but more weakly (circled
coefficient). \textbf{B.} In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in
locations without trees makes their occurrence vectors positively
correlated (circled). \textbf{C.} Controlling for trees with a
conditional (all-else-equal) approach such as a partial covariance or a
Markov network leads to correct identification of the negative
shrub-shrub interaction (circled).

\end{flushleft}
\end{spacing}

>>>>

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that is capable of
outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can
predict the direct and indirect consequences of any possible species
interaction matrix. Interactions in these models can be estimated from
observed co-occurrence rates via maximum likelihood, controlling for
indirect effects. Using simulated landscapes with known pairwise
interaction strengths, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods,
correctly isolating direct interactions between species pairs even when
indirect interactions or abiotic factors largely overpowered them. Two
computationally efficient approximations, based on controlling for
indirect effects with linear or generalized linear models, also
performed well. Indirect effects reliably caused a common null modeling
approach to produce incorrect inferences, however.

@@

\noindent
To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Connor et al. 2013), existing methods
for detecting competition's effects on community structure are
unreliable (Gotelli and Ulrich 2009). In particular, species' effects on
one another can become lost in a web of indirect effects. For example,
the competitive interaction between the two shrub species in Figure 1A
is obscured by their shared tendency to occur in unshaded areas (Figure
1B). While ecologists have long known that indirect effects can
overwhelm direct ones (Levine 1976), most methods for drawing inferences
from co-occurrence data do not control for these effects (e.g. Diamond
1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013, Pollock
et al. 2014). As a result, ecologists do not have tools that allow them
to isolate direct interactions from indirect effects.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most methods assume), it does still leave a
signal in the data (Figure 1C). After controlling for the presence of
the tree species (e.g.~by splitting the data set into shaded and
unshaded sites or by using a model that estimates conditional
relationships amongs species), the two shrubs do have a negative
association, with fewer co-occurrences in unshaded areas than would be
expected if they occurred independently of one another.

Following Azaele et al. (2010), this paper shows that Markov networks
(undirected graphical models also known as Markov random fields; Murphy
2012) can provide a framework for understanding the landscape-level
consequences of pairwise species interactions, and for estimating them
from observed presence-absence matrices. Markov networks have been used
in diverse scientific fields in similar contexts for decades, from
physics (where nearby particles interact magnetically; Cipra 1987) to
spatial statistics (where adjacent grid cells have correlated values;
Harris 1974, Gelfand et al. 2005). While community ecologists explored
some related approaches in the 1980's (Whittam and Siegel-Causey 1981),
they used severe approximations that led to unintelligible results (e.g.
``probabilities'' greater than one; Gilpin and Diamond 1982).

Below, I demonstrate Markov networks' ability to produce exact
predictions about the direct and indirect consequences of an interaction
matrix, and also to make inferences about the interaction matrix based
on co-occurrence rates. Using simulated data sets where the ``true''
interactions are known, I compare this approach with several existing
methods. Finally, I discuss opportunities for extending the approach
presented here to other problems in community ecology, e.g.~quantifying
the overall effect of species interactions on occurrence rates
(Roughgarden 1983) and disentangling the effects of biotic versus
abiotic interactions on species composition (Pollock et al. 2014).

@@

\noindent where the second sum is over all \(\frac{1}{2}n(n-1)\) pairs
of \(n\) species. In this model, \(\alpha_{i}\) is an intercept term
determining the amount that the presence of species \(i\) contributes to
the log-probability of \(\vec{y}\); it directly controls the prevalence
of species \(i\). Similarly, \(\beta_{ij}\) is the amount that the
co-occurrence of species \(i\) and species \(j\) contributes to the
log-probability; it controls the conditional relationship between two
species, i.e.~the probability that they will be found together, after
controlling for the other species in the network (Figure 2A, Figure 2B).
For example, if \(\beta_{ij} = +2\), then each species' odds of
occurrence would be \(e^2\) times higher when the other one is present
(as compared with otherwise equivalent sites). The relative probability
of a presence-absence vector increases when positively-associated
species co-occur and decreases when negatively-associated species
co-occur. As a result, the model tends---all else equal---to produce
assemblages where many positively-associated species pairs co-occur and
few negatively-associated pairs do (just as an ecologist might expect).
When all else is \emph{not} equal (e.g.~Figure 1, where the presence of
one competitor is associated with release from another competitor), then
predicting species' overall co-occurrence rates would require
normalizing the relative probabilities of possible \(\vec{y}\) vectors
so that they sum to one (Figure 2B).

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When Equation 1 can be
normalized, one can find exact maximum likelihood estimates for
\(\alpha\) and \(\beta\) using numerical optimization. Fully-observed
networks like the ones considered here have unimodal likelihood surfaces
(Murphy 2012), so these estimates will represent the global optimum. I
used the rosalia package (Harris 2015a) for the R programming language
(R Core Team 2015) to calculate \(p(\vec{y}; \alpha, \beta)\) and its
gradients (Murphy 2012); the package passes these functions to the
``BFGS'' method in R's general-purpose optimizer, which then estimates
the Markov network parameters.

@@

I then simulated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model). The simplest set of
simulated landscapes were generated with Gibbs samples from Equation 1,
and provided a test of methods' ability to isolate direct species-level
interactions from a web of indirect ones. For each replicate, I randomly
drew the ``true'' \(\beta\) coefficient magnitudes an exponential
distribution with rate 1 so that most species pairs interacted
negligibly but a few interactions were strong enough to propagate
through the newtwork. I randomly assigned 25\% of the interactions to be
positive; the remainder were negative.

The next set of landscapes provided a way to assess each method's
ability to identify direct interactions in the presence of environmental
heterogeneity. Here, the \(\beta\) coefficients were calculated as
above, but each species' \(\alpha\) value depended linearly on two
normally-distributed environmental factors, which were drawn from
independent Gaussians for each site. Once the local \(\alpha\) values
were calculated, independent Gibbs samplers determined the species
composition for each site based on Equation 1.

In the third set of landscapes, I simulated each species' abundance
(instead of just their presence/absence), and interactions between
species occurred on a per-capita basis. To avoid runaway mutualisms
leading to infinite abundance, all interaction coefficients were
negative in these simulations. If a method performs well on these
landscapes, it would indicate some robustness to the biolocical
mechanisms underlying species' interactiosn with one another.

@@

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships. This approach, which is closely related to
linear regression, is common in molecular biology (Friedman et al.
2008), but is rare in ecology (see Albrecht and Gotelli (2001) and
Faisal et al. (2010) for two exceptions). In the context of non-Gaussian
data, the partial correlation can be thought of as a computationally
efficient approximation to the full Markov network model (Loh and
Wainwright 2013). Partial correlations are undefined for landscapes with
perfectly-correlated species pairs, so I used the regularized estimate
provided by the corpcor package's \texttt{pcor.shrink} function with the
default settings (Schäfer et al. 2014).

@@

The next method used the Pairs software described in Gotelli and Ulrich
(2009). This program simulating new landscapes from a null model that
retains the row and column sums of the original matrix (Strong et al.
1984) and calculates \(Z\)-scores to summarize a species pair's
deviation from this null.

@@

For the larger landscapes, I evaluated the relationship between each
method's estimates and the ``true'' interaction strengths. To ensure
that the different test statistics (e.g.~correlations versus \(Z\)
scores) were on a common scale, I rescaled them using linear regression
through the origin. I then calculated the proportion of variance
explained for different combinations of model type and landscape size
(compared with a baseline model that assumed all interaction strengths
to be zero).

For the null model and the Markov network, the probability of rejecting
the null hypothesis of zero interaction was estimated across a range of
``true'' interaction strengths using a generalized additive model with
binomial responses (reject versus not; Appendix 4). The Type I error
rate was defined as this model's estimated probability of rejecting the
null hypothesis when the ``true'' interaction was zero.

@@

\noindent
\textbf{Twenty species.} In general, each model's performance was
highest for large landscapes with simple assembly rules and no
environmental heterogeneity (Figure 3). Despite some variability across
contexts, the rank ordering across methods was very consistent. In
particular, the four methods that controlled for indirect effects (the
Markov network, the generalized linear models, and the two partial
correlation-based methods) always matched or outperformed those that did
not. The Markov network consistently performed best of all. As
anticipated by Lee and Hastie (2012), generalized linear models closely
approximated the Markov network estimates (Figure 4A), especially when
the data sets were very large (Figure 3). As reviewed in Gotelli and
Ulrich (2009), however, most analyses in this field of ecology involve
fewer than 50 sites; in this context, the gap between the methods was
larger. As shown in Appendix 4, the standard errors associated with the
estimates in Figure 3 are small (less than 0.01), so the differences
among methods should not be attributed to sampling error.

Of the methods that did not control for indirect effects, Figure 3 shows
that simple correlation coefficients provided a more reliable indicator
of species' true interaction strengths than either the joint species
distribution model (BayesComm) or the null model (Pairs). 95\% of the
variance the Pairs test statistic was explained by correlation
coefficients (controlling for landscape size; Figure 4B); much of the
remaining variance is due to sampling error.

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, where the interaction coefficients are
easiest to interpret). The Markov network's Type I error rate depended
on the simulation type: {[}{[}0.02{]}{]} for simulations that matched
the model's assumptions, versus {[}{[}0.14{]}{]} for simulations that
included environmental heterogeneity (see Appendix 4 for confidence
interval coverage across a range of \(\beta_{ij}\) values). The null
model's Type I error rates were {[}{[}0.30{]}{]} and {[}{[}0.50{]}{]}
for the constant and heterogeneous landscapes, respectively---far higher
than the nominal 0.05 rate. Figure 4C shows, across a range of true
interaction strengths, the probability that the null model or the Markov
network will predict the wrong sign of the interaction with 95\%
confidence. The null model makes such errors more than {[}{[}5{]}{]}
times as often as the Markov network, even though it only rejects the
null hypothesis {[}{[}twice{]}{]} as often overall (Appendix 4). The
Markov network's errors were also more concentrated around 0, so it
never misclassified strong interactions like the null model did.

@@

\setlength{\parindent}{-1em} \setlength{\leftskip}{1em}
\setlength{\parskip}{0pt}

\textbf{Figure 1: A.} A small network of three competing species. The
tree (top) tends not to co-occur with either of the two shrub species,
as indicated by the strongly negative coefficient linking them. The two
shrub species also compete with one another, but more weakly (circled
coefficient). \textbf{B.} In spite of the competitive interactions
between the two shrub species, their shared tendency to occur in
locations without trees makes their occurrence vectors positively
correlated (circled). \textbf{C.} Controlling for trees with a
conditional (all-else-equal) approach such as a partial covariance or a
Markov network leads to correct identification of the negative
shrub-shrub interaction (circled). See Appendix 1 and the results for
``three species'' for more details.

@@

\hyperdef{}{ref-xiaoux5fstrongux5f2015}{\label{ref-xiaoux5fstrongux5f2015}}
Xiao, X., D. J. McGlinn, and E. P. White. 2015. A strong test of the
Maximum Entropy Theory of Ecology. The American Naturalist 185:E70--E80.
\end{flushleft}
\end{spacing}
